The Economics of Wildfire Aviation

### 2023-09-07 manuscript/manuscript.qmd
penultimate draft for first submission

---
title: "The Economics of Wildfire Aviation"
format: 
  docx:
    reference-doc: custom-reference-doc.docx
    toc: false
    number-sections: true
bibliography: references.bib
csl: journal-of-agricultural-and-resource-economics.csl
---

@@

<!-- objectives  -->

@@

<!-- aviation is an important response asset but expensive -->

@@

### Maybe material

-   more flexible CWN contracts: Current records suggest that CWN contracts are executed for 90 days. The federal firefighting agencies could benefit from shorter and more flexible contracts. If there were no transactions cost[^6], the ideal arrangement would allow the firefighting agency to use (and pay for) the aircraft only when they need it.

[^6]: By transactions cost, we mean the time and effort to arrange, verify, and execute the contract, the moving of aircraft, and any other logistical issues involved in using the aircraft.

For example, the default CWN contract could be one week with a requirement that the federal agency renew or release the aircraft before

-   sunk cost fallacy: The structure of the procurement model may influence how aviation resources are used. The daily availability price can range from \$30k to \$50k per day while the hourly flight cost ranges from \$xxk to \$xxk. Those involved in allocating aviation resources may view unused aircraft as wasteful since the daily availability rate is paid regardless of use.

-   The schedule of items suggests that there are relatively few private entities that own aircraft. Procurement auctions tend to be more economically efficient when there are more bidders competing for the limited contract. Auction theory suggests that the competition between the bidders over a limited number of contracts will incentivize them to reveal their true willingness to accept, which should equal their marginal cost of provision. Moreover, the auction efficiency assumes that each bidder's cost of provision is unknown to the other bidders. When the number of bidders is small and there is ample opportunity for the bidders to communicate in advance of submitting bids, it is more likely that the bidders will bid above their marginal cost of provision. Moreover, the aircraft owners repeatedly participate in this annual procurement process allowing them to learn about their competitors and the government's willingness to pay. In principle, these forces may lead to higher procurement costs. However, these auction inefficiencies may be unavoidable since the ownership and maintenance of large aircraft is capital intensive, which serves as a barrier to new entrants in the market.[^7]

[^7]: The McDonnell Douglas DC 10, one of the largest active airtankers, is currently listed for sale for \$2.75 million (https://myemail.constantcontact.com/McDonnell-Douglas-DC-10-30-Freighter-For-Sale--Price-Reduction-.html?soid=1102673403687&aid=zTC4sV-oTNI, accessed September 9, 2023).

Moreover, the federal agencies are not the only customers of companies that own firefighting aircraft. Indeed, these companies increasingly sell their services to other countries (e.g., Canada and Australia) as well as US state agencies. For example, the Colorado Division of Fire Prevention and Control has signed exclusive use contracts for a BA-146 airtanker provided by Neptune Aviation [@gabbert2021].

@@

>>>>

---
#title: "The Economics of Wildfire Aviation"
format: 
  docx:
    reference-doc: custom-reference-doc.docx
    toc: false
    number-sections: true
bibliography: references.bib
csl: journal-of-agricultural-and-resource-economics.csl
---

## Abstract

We examine two economic aspects of aviation used for wildland firefighting in the United States. First, we analyze the procurement of aviation resources by federal agencies, and how the structure of contracts influences suppression costs. Second, we highlight the lack of empirical evidence on the effectiveness of aviation, and provide a research design to quantify effectiveness and the benefits of aviation use. Our analysis can inform tactical operations on individual wildfire incidents and help federal agencies develop season-long resource utilization strategies.
[^ack]

[^ack]: Acknowledgements: We acknowledge support from the United States Forest Service (agreement #21-CS-11221636-151). We are thankful for the research assistance from Kyra Adamson.

@@

@@

@@

@@

## Figures
